---
title: "Biodiversity in National Parks"
author: "GSD"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    number_sections: yes
    code_folding: show
    theme: sandstone
    highlight: textmate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


#Loading required libraries,
```{r}
library(tidyverse)
library(knitr)
library(data.table)
library(formattable)
parks=read.csv("parks.csv",header=TRUE,stringsAsFactors = FALSE)
species=read.csv("species.csv",header=TRUE,stringsAsFactors = FALSE)
glimpse(parks)
glimpse(species)
```

# List of parks by acres

There are `r length(unique(parks$Park.Name))` parks.


```{r,fig.width=8,fig.height=8}

parks %>% arrange(-Acres) %>% ggplot(aes(factor(Park.Name,levels=Park.Name),Acres,fill=Park.Name))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family="verdana",hjust=0.5),legend.position = "none")+labs(x="Park Name",y="Acres")+ggtitle("Parks and the Acres")+coord_flip()
```

#Count of each category,Abundance level,Nativity,Family,Parks with highest species count,Conservation status,Seasonality

## By Category

There are `r length(unique(species$Category))` categories in total.

```{r}
temp=species %>% group_by(Category) %>% summarise(Category.Count=n()) %>% arrange(-Category.Count)
ggplot(temp,aes(factor(Category,levels=Category),Category.Count,fill=Category))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family="verdana",hjust=0.5),legend.position = "none")+labs(x="Category",y="Total Species")+ggtitle("Total species in each category")
```


## By Abundance level

```{r}
length(unique(species$Abundance))
temp=species %>% group_by(Abundance) %>% summarise(Abundance.Count=n()) %>% arrange(-Abundance.Count)
temp[temp$Abundance=="",1]="Unspecified"
ggplot(temp,aes(factor(Abundance,levels=Abundance),Abundance.Count,fill=Abundance))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family = "Verdana",hjust=0.5),legend.position = "none")+labs(x="Abundance",y="Total Species")+ggtitle("Total species in each Abundance level")

```

As seen from the bar chart,most of the category of species are not recorded into abundance level and it is impossible to draw conclusions about the abundance level of each species.

## Nativity:

```{r}
length(unique(species$Nativeness))
temp=species %>% group_by(Nativeness) %>% summarise(Nativeness.Count=n()) %>% arrange(-Nativeness.Count)
temp[temp$Nativeness=="",1]="Unspecified"
ggplot(temp,aes(factor(Nativeness,levels=Nativeness),Nativeness.Count,fill=Nativeness))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family = "Verdana",hjust=0.5),legend.position = "none")+labs(x="Nativeness",y="Total Species")+ggtitle("Total species in each Nativeness level")

```

Native species are found in abundance in the parks with over 65000 species comming under them.There are about more than 11,000 non native species.

## Family 

```{r}
length(unique(species$Family))
temp=species %>% group_by(Family) %>% summarise(Family.Count=n()) %>% arrange(-Family.Count)
temp[temp$Family=="",1]="Unspecified"
ggplot(head(temp,20),aes(factor(Family,levels=Family),Family.Count,fill=Family))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family = "Verdana",hjust=0.5),legend.position = "none",plot.subtitle=element_text(hjust=0.5))+labs(x="Family",y="Total Species")+ggtitle("Total species in each Family",subtitle = "Top 20 Families")+coord_flip()

```

##Conservation Status

```{r}
length(unique(species$Conservation.Status))
temp=species %>% group_by(Conservation.Status) %>% summarise(Count=n()) %>% arrange(-Count)
temp[temp$Conservation.Status=="",1]="Unspecified"
ggplot(temp,aes(factor(Conservation.Status,levels=Conservation.Status),Count,fill=Conservation.Status))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family = "Verdana",hjust=0.5),legend.position = "none",plot.subtitle=element_text(hjust=0.5))+labs(x="Conservation Status",y="Total Species")+ggtitle("Conservation level of each species")+coord_cartesian(ylim=c(0,12000),expand=TRUE)+scale_y_continuous(breaks=seq(0,12000,1000))
```

Most of the species are categoried as unspecified .Above 3000 species are species of concern while less than 500 of them are endangered.

Now,let us examine in depth,



#Which category of species is found in abundance?

In total there are 56 parks as evident from the parks data.

```{r}
species %>% filter(Abundance=="Abundant") %>% group_by(Category) %>% summarise(abundance.count=n()) %>% arrange(-abundance.count) %>% ggplot(aes(factor(Category,levels=Category),abundance.count,fill=Category))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family="verdana",hjust=0.5),legend.position = "none")+labs(x="Category",y="Count")+ggtitle("Category of species found in Abundance in 56 parks")+coord_flip()
```

#Which category of species is endangered?

```{r}
temp=species %>% filter(Conservation.Status=="Endangered") %>% group_by(Category) %>% summarise(count=n()) %>% arrange(-count)
ggplot(temp,aes(factor(Category,levels=Category),count,fill=Category))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family="verdana",hjust=0.5),legend.position = "none")+labs(x="Category",y="Count")+ggtitle("Category of species which are endangered")

```

Vascular plants are endangered the most followed by Mammals.

#Which park has maximum Vascular plants?

```{r}
temp=species %>% filter(Category=="Vascular Plant") %>% group_by(Park.Name) %>% summarise(count=n()) %>% arrange(-count) 
ggplot(head(temp,10),aes(factor(Park.Name,levels=Park.Name),count,fill=Park.Name))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family="verdana",hjust=0.5),legend.position = "none")+labs(x="Category",y="Count")+ggtitle("Top 10 Parks having maximum Vascular Plants")
```

_Shenandoah National Park,Death Valley National Park,Redwood National Park,Great Smoky Mountains National Park_ has more than 2000 Vascular plants.

#Which park has maximum endagered Vascular plants ?


```{r,fig.width=14,fig.height=17}
temp=species %>% filter((Conservation.Status=="Endangered") & (Category=="Vascular Plant")) 
ggplot(temp,aes(Park.Name,fill=Scientific.Name))+geom_bar(stat="count")+theme(axis.text.x = element_text(angle=90,vjust=0.5),plot.title=element_text(family="verdana",hjust=0.5),legend.position = "bottom")+labs(x="Park",y="Species Count")+ggtitle("Parks where endangered species are found")+coord_flip()

```

_Hawaii Volcanoes National Park ,Haleakala National Park , Channel Island National Park ,Death Valley National Park_ has maximum number of endangered vascular plants.

#Which species is found in most parks?

```{r}
temp=species %>% group_by(Common.Names) %>% summarise(Count=n()) %>% arrange(-Count)
temp[temp$Common.Names=="None",1:2]=NA
temp=na.omit(temp)
summary(temp)
```


Here 75 % of the species are found in less than 5 parks.Therefore each park has distinct species of its own.




